March Madness Basketball Turniment

### 2021-04-15 slides/slides.Rmd
add finished slides

# Backround

@@

- First tournament was in 1939 only 8 teams

@@

- Now, its restricted to the top 64 NCAA Division I Teams

@@

- Only one team has ever gone undefeated thought a season and the tournament

@@

- 16.9 Million Viewers

@@

- MSU won in 1979 and 2000

@@

- When the turiment starts we form small groups and gamble on the outcomes. Usually by selecting winners and loser from the backet until we have a completely filled out turniment

@@

- Nobody has ever made a perfect bracket and there are $2^{63}$ possible outcomes.

@@

---
class: center, middle
# Zoomed in

@@

# Using basketball seasion data to predict this year's turniment outcomes requires us to:

### Design predive models to make out of sample predictions

@@

### Derrive a scoring metric

@@

### Evaluate and tune the model

@@

### We made several models, In the interstst of time we are going to discuss the logistic because its the most straight-forward

@@

histogram <- function(x, t){
  ggplot(data = hists, aes(x)) +
    geom_histogram(bins = 55, color = "white", fill = "black") +
    theme_light(base_size = 25) +
    labs(title = paste0("", t), x = "", y = "") +
    NULL

}

a <- histogram(merged$x3fg,       t = "Three-Points Goals")
b <- histogram(merged$fg_percent, t = "Percentage of Goals Scored")
c <- histogram(merged$rpg,        t = "Rebounds Per Game")
d <- histogram(merged$st,         t = "Steals Per Game")
e <- histogram(merged$to,         t = "Turnovers Per Game")
f <- histogram(merged$bkpg,       t = "Blocks Per game")
# (a + b + c) / (d + e + f)
(a + b) / (c + d) /(e + f)
```

@@

- Normal-ish distributions

- No catastrophic skews or heavy tails

@@

- Ideally we would have had steaks and turn-overs per game rather than season total. Not all played the same game

@@

@@

```{r}

table_output(obj = logistic_fit, rows = 2:3)
```

@@

```{r}
table_output(obj = logistic_fit, rows = 4:5)
```

@@

```{r}
table_output(obj = logistic_fit, rows = 6:7)
```

@@

```{r}
table_output(obj = logistic_fit, rows = 8:9)
```

@@

```{r}
table_output(obj = logistic_fit, rows = 10:11)
```

@@

```{r}
table_output(obj = logistic_fit, rows = 12:13)
```

@@

```{r}
table_output(obj = logistic_fit, rows = 14:15)
```

@@

- Our goal Goal is to weight correct prediction higher when it assigns a higher probability. Conversely a lower score to an incorrect prediction.

@@

- $E(Model\ Score) = 0$ if our predictions are no better than randomly guessing.

@@

@@

```{r}
logistic_model_score %>% 
  kable(align = "r") %>% 
  kable_styling(font_size = 15) %>% 
  column_spec(c(5), color =  ifelse(logistic_model_score$Score > 0, "white", "black"), 
              background = ifelse(logistic_model_score$Score > 0, "#025cba", "#ff3700"))
```

@@

```{r}
t2 <- perform %>% 
  summarise(totalscore = sum(score),
            se = sd(score), 
            t = totalscore/se,
              p = pt(t, nrow(perform)-1, lower.tail = F)) %>% 
  t()

@@

t1 <- rbind(score, c("Total", sum(score$totalscore))) 
```

```{r}
knitr::kable(list(t1, t2)) %>% 
  kable_styling(font_size = 10)

@@

@@

### Logistic Regression performs well. Given measured statists.

### However out model is independent of the pervious game which would aplify shortcommings.

>>>>

# Backround on March Madness Tournament

@@

- First hosted in 1939 with 8 participating teams.

@@

- Now, restricted to the top 64 qualifying NCAA division I teams.

@@

- Currently 16.9 million viewers.

@@

- Michigan State won in 1979 and 2000.

@@

- Offices, organizations, and friend groups make and submit a total of 40 Million predictions each year.

@@

- At the beginning small groups form and gamble on the outcomes by selecting winners and losers.

@@

- Nobody has ever made a perfect bracket prediction and there are $2^{63}$ possible outcomes.

@@

![](images/results.jpg)

class: center, middle
# Zoomed in

@@

## We use basketball season game scores and statistics to predict this year's tournament outcome.

@@

## We need to derive a scoring metric.

@@

## Evaluate and tune the model.

@@

## Compare the model's performance.

## We made several models, in the interest of time we are going to discuss the logistic because its the most straight-forward.

@@

histogram <- function(x, t){
  ggplot(data = hists, aes(x)) +
    geom_histogram(bins = 55, color = "white", fill = "#025cba") +
    theme_light(base_size = 25) +
    labs(title = paste0("", t), x = "", y = "") +
    NULL
}

a <- histogram(merged$x3fg,       t = "Three-Points Goals")
b <- histogram(merged$fg_percent, t = "Percentage of Goals Scored")
c <- histogram(merged$rpg,        t = "Rebounds Per Game")
d <- histogram(merged$st,         t = "Steals Per Game")
e <- histogram(merged$to,         t = "Turnovers Per Game")
f <- histogram(merged$bkpg,       t = "Blocks Per Game")
# (a + b + c) / (d + e + f)
(a + b) / (c + d) /(e + f)
```

@@

- Normal-ish distributions.

- No catastrophic skews or heavy tails.

@@

- Ideally we would have had steaks and turn-overs per game rather than season total. Not all played the same game.

@@

lrt_stat <- 2*(as.numeric(logLik(logistic_fit)) -
  as.numeric(logLik(logistic_fit_null)))

pval_lrt <- pchisq(lrt_stat, df = 1, lower.tail = FALSE)

dev <- deviance(logistic_fit)

pval_dev <- pchisq(dev, df.residual(logistic_fit), 
       lower.tail = FALSE)

tests <- cbind(
LRT = c(lrt_stat, pval_lrt),
DEViance = c(dev, pval_dev)
)

rownames(tests) <- c("Stat", "P Value")
lrt_dev <- kable(tests) %>% 
  kable_styling(font_size = 10)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 2:3)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 4:5)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 6:7)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 8:9)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 10:11)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 12:13)
lrt_dev
```

@@

```{r}
table_output(obj = logistic_fit, rows = 14:15)
lrt_dev
```

@@

- Our goal is to weight correct prediction higher when it assigns a higher probability. Conversely, a lower score to an incorrect prediction.

@@

- If our predictions are no better than randomly guessing $E(Model\ Score) = 0$.

@@

Incorrect Predictions

```{r}
logistic_model_score %>% 
  kable(align = "r") %>% 
  kable_styling(font_size = 15) %>% 
  column_spec(c(1:5), color =  ifelse(logistic_model_score$Score > 0, "black", "white"), 
              background = ifelse(logistic_model_score$Score > 0, "white", "#ff3700"))
```

@@

@@

```{r}
t2 <- perform %>% 
  summarise(totalscore = sum(score),
            se = sd(score), 
            t = totalscore/se,
              p = pt(t, nrow(perform)-1, lower.tail = F)) %>% 
  t() %>% 
  tibble() %>% 
  mutate_if(is.numeric, round, 2)

rownames(t2) <- c("Total Score", "Standard Error", "T-Statistic", "P-value")
colnames(t2) <- NULL

@@

t1 <- rbind(score, c("Total", sum(score$totalscore))) 
colnames(t1) <- c("Result", "Score")

@@

```{r}
knitr::kable(list(t1, t2)) %>% 
  kable_styling(font_size = 10)
```
---
class: center
Model Performance

```{r}
a <- perform %>% 
  ggplot(aes(as.factor(outcome), fill = outcome)) + 
  geom_bar(alpha = 0.7) + 
  labs(title = "#2 Logistic Model", x = "", y = "") + 
  scale_fill_manual(values = c("#025cba", "#ff3700")) +
  geom_text(size=10, color = "grey80", stat = 'count',
            aes(label =..count.., hjust = 1.5)) + 
  coord_flip()
```

```{r}
posson_model_score <- readRDS(here::here("cache","posson_model_score.RDS"))

b <- posson_model_score %>% 
  drop_na() %>% 
  ggplot(aes(as.factor(outcome), fill = outcome)) + 
  geom_bar(alpha = 0.7) + 
  labs(title = "#1 Poisson Model", x = "", y = "") + 
  scale_fill_manual(values = c("#025cba", "#ff3700")) +
  geom_text(size=10, color = "grey80", stat = 'count',
            aes(label =..count.., hjust = 1.5)) + 
  coord_flip()
```

```{r}
multinomeal_model_score <- readRDS(here::here("cache", "multinomeal_model_score.RDS"))

c <- multinomeal_model_score %>% 
  drop_na() %>% 
  ggplot(aes(as.factor(outcome), fill = outcome)) + 
  geom_bar(alpha = 0.7) + 
  labs(title = "#3 Multinomeal Model", x = "", y = "") + 
  scale_fill_manual(values = c("#025cba", "#ff3700")) +
  geom_text(size=10, color = "grey80", stat = 'count',
            aes(label =..count.., hjust = 1.5)) + 
  coord_flip()
```

```{r fig.height=11}
b/a/c
```

@@

### Both Logistic and Poisson Regression perform well.

### However, we are scoring the model based on independent predictions. While real-world predictions are made dependent on the previous round.

### These predictions are not wildly different than betting markets and seeds because they use similar statistics to derive their predictions.
